The Role of Variability in Learning Generalization: A Computational Modeling Approach

### 2024-05-28 Presentation/_hold_slides.qmd
sync again

>>>>

### Final Conclusion Slide

**Title: Conclusion**

**Bullet Points:**
- **Empirical Findings**:
  - HTT: Varied training led to superior performance.
  - HTW: Varied training led to poorer performance.
  - Differences attributed to task characteristics and complexity.

- **Modeling Insights**:
  - HTT: IGAS model highlighted the importance of similarity between training and testing conditions.
  - HTW: EXAM model provided the best account of testing data, especially for the constant group.

- **Implications**:
  - Task-specific characteristics are crucial in determining the benefits of varied training.
  - Combining empirical research with computational modeling enhances understanding of learning and generalization.

**Script:**

"In summary, our research provided significant insights into the effects of training variability on learning and generalization in different tasks. In the HTT project, we found that varied training led to superior performance, consistent with the benefits of variability hypothesis. In contrast, the HTW project demonstrated that varied training could be detrimental, particularly in simpler extrapolation tasks.

These opposing patterns highlight the importance of task characteristics and complexity in determining the benefits of training variability. The HTT task's complexity and perceptual salience of varied conditions likely contributed to the positive effects, while the simpler, less perceptually salient HTW task showed negative effects.

From a modeling perspective, the IGAS model in HTT emphasized the role of similarity between training and testing conditions, showing that controlling for similarity is crucial for understanding training outcomes. The EXAM model in HTW provided a robust account of the data, particularly for the constant training group, suggesting that linear extrapolation mechanisms and prior knowledge can influence performance in simpler tasks.

Overall, our findings underscore the necessity of considering task-specific characteristics when designing training programs. The combination of empirical research and computational modeling proved invaluable in uncovering the cognitive mechanisms underlying learning and generalization. Future research should continue to explore the interplay between task demands, training manipulations, and individual differences to optimize educational and training outcomes across various domains. Thank you."

### 2024-05-28 Presentation/slides.qmd
sync again

@@

- Variation during training linked to improved transfer in numerous domains 
  - sometimes alternatively termed diversity, heterogenity, numerosity etc.

- What does variability mean in the context of learning interventions?
  - How spread out examples are in the task space
  - The number of unique items/problems experienced
  - exposure to wider array of contexts/background conditions

@@

@@

@@

@@

::: notes
Variety of theoretical explanations
Schema theory in paticular has been extremely influential 
:::

@@

@@

@@

>>>>

::: notes
thank you all for joining in my defense

Learning is often more specific than we'd like. There has been longstanding scientific interest in trying to understand what factors we might
be able to control that might have beneficial effects on generalization and transfer.

One such factor, and the topic of my dissertation - is the influence of variability on generalization

Over the next 45 minutes, I will walk through the specific empirical studies and computational modeling work

@@

- Variation during training linked to improved transfer in numerous domains 
  - Category learning, perceptual learning, education, visuomotor learning
  - sometimes alternatively termed diversity, heterogenity, numerosity etc.

::: {.fragment}
- What does variability mean in the context of learning interventions?
  - How spread out examples are in the task space
  - **The number of unique items/problems experienced**
  - exposure to wider array of contexts/background conditions
:::

- The influence of variability has been studied across many different domains
- although sometimes referred

variability can refer to many different

@@

::: notes
higher or lower variability can of course vary greatly between domains

One common manipulation in visuomotor learning is to have a group train with the lowest possible variability - constant

This is a fairly common pattern, at least to my reading. But....,

@@

::: notes
Plenty of discrepancy in results, sometimes even for very similar tasks. 
- and lots of work remains to unravel when exactly variability is or isn't helpful
:::

@@

::: notes
- Dissertations consists of 2 primary projects. 
- Each within their own subdomain of visuomotor learning
- and each employing a distinct type of computational modeling
:::

@@

::: notes
- Variety of theoretical explanations have been proposed for the commonly observed effects
- In visuomotor learning and most relevant for my work, Schema theory in paticular has been extremely influential, and seems to have inspired hundreds of studies
:::

@@

::: notes
- Schema theory assumes that learners encode abstractions, or parameters, that enable generalization, and that variation is parituiclarly helpful
:::

@@

::: notes
- To compare accuracy between groups in the testing stage, we fit a Bayesian mixed effects model predicting deviation from the target band as a function of training condition (varied vs. constant) and band type (trained vs. extrapolation), with random intercepts for participants and bands. The model results are shown in Table 3. The main effect of training condition was not significant (
 The extrapolation testing items had a significantly greater deviation than the training bands (
  Most importantly, the interaction between training condition and band type was significant (
 As shown in Figure 5, the varied group had disproportionately larger deviations compared to the constant group in the extrapolation bands.

 - Finally, to assess the ability of both conditions to discriminate between velocity bands, we fit a model predicting velocity as a function of training condition and velocity band, with random intercepts and random slopes for each participant.

- on. Most relevant to the issue of discrimination is the coefficient on the Band predictor (
 Although the median slope does fall underneath the ideal of value of 1, the fact that the 95% credible interval does not contain 0 provides strong evidence that participants exhibited some discrimination between bands. The significant negative estimate for the interaction between slope and condition (
 suggests that the discrimination was modulated by training condition, with the varied participants showing less sensitivity between bands than the constant condition 
:::

@@

::: notes
 The analysis of testing accuracy examined deviations from the target band as influenced by training condition (Varied vs. Constant) and band type (training vs. extrapolation bands). The results, summarized in Table 6, reveal no significant main effect of training condition (
 However, the interaction between training condition and band type was significant  with the varied group showing disproportionately larger deviations compared to the constant group on the extrapolation bands
